Hire Ed: Job Market Dynamics for Tenure-Track Faculty Positions in Archaeology

### 2025-05-12 analysis/paper/article.qmd
Minor edits to text

Junior scholars with faculty ambitions are thus stuck in a double-bind, forced to funnel their energy into seemingly endless applications for which they receive little to no feedback and which remain invisible on their CVs. As Dennis et al. underscore, early career researchers are "supplicants": "There is no way for applicants to point out that the job ads are taking too much time out of the scholarly community’s collective time bank" [@dennisWorstAnthroJob2022, pp. 1]. Indeed, acknowledging the new reality that applying for academic positions is in and of itself a full-time job, a growing number of anthropology graduate programmes are offering professionalization courses that cover the ins and outs of everything from crafting strong cover letters and CVs, to performing well in long-list interviews and preparing teaching demonstrations. These courses can provide valuable training in the hidden curriculum of the academy, but the fact that job applications are so complicated that they require training to navigate suggests more disciplinary attention should be paid to the "market" as a historically contingent set of cultural practices. As Rabinow maintains, "if we want ethical considerations to play a central role in the articulation of truth and power—and I think we do—then bringing such considerations into view is the necessary first step towards recognizing who we are today and setting out on the road to a better place" [-@rabinow1992, pp. 71].

@@

PhD graduates who do not go directly into a tenure-track faculty position after graduation often go into a series of short-term appointments as part-time or limited contract instructors, also known as adjunct or contingent faculty. The rising number of contingent faculty has long been a concern in US higher education [@Trevithick2010]. Data from the American Association of University Professors' 2021–22 faculty survey indicate that more than 60% of faculty positions in US universities were held by non-tenure-track full time or part time contingent faculty members [@colby2022annual]. Contingent faculty positions are precarious because they provide low or no health and retirement benefits, limited opportunities for professional development, and lower salaries relative to tenure-track positions. PhD graduates who have a sequence of short-term academic positions are often disadvantaged financially due to low compensation combined with expensive frequent relocation, socially due to isolation from family and community, and professionally due to being ineligible for many decision-making roles at the universities at which they work [@platzerAcademicPrecarityAmerican2018].

Success in tenure-track job applications in archaeology is strongly determined by where applicants get their PhDs. Data from the 2014–2015 AnthroGuide publication of the American Anthropological Association shows that just ten out of over 100 US graduate programs produced over 30% of the graduates hired into tenure-track faculty positions [@speakman2018choosing]. Top-ranking programs are placing fewer than one in three graduates in tenure-track jobs [@mackieMarketShareAccounting2023]. Similarly, a network analysis of 1,918 faculty holding tenured or tenure-track positions at PhD-granting anthropology programs in the US in 2015 (including 506 archaeologists) found that the just fifteen graduate programs produced 53 percent of tenured and tenure-track positions [@kawaSocialNetworkUS2019]. This network analysis showed that programs with large endowments and faculty who hold prestigious awards and are widely cited by other scholars, produce the majority of tenured and tenure-track faculty.

Hiring bias predicated on the prestige of specific institutions and programs is not unique to anthropology. Targeted studies of sociology, communication, operations research, and industrial systems engineering show similar dynamics [@barnett2010; @castillo; @feeley2021; @nevin2019]. Broader analyses comparing computer science, business, and history likewise comparable patterns among disparate disciplines across the humanities, social sciences, and STEM fields; "across disciplines, prestige hierarchies make the most accurate predictions of faculty placement" [@clauset2015, pp. 4]. The most recent comprehensive meta-analysis of faculty placement dynamics in the United States, which examined employment and doctoral education of all tenure-track faculty at PhD-granting universities, underscored the outsize impact of institutional prestige: 80% of faculty trained in the US came from just 20% of institutions [@wapmanQuantifyingHierarchyDynamics2022a]. Within anthropology, however, such studies draw attention to the paradox of a discipline committed to fighting social inequalities continuing to reproduce its own systematic inequalities through hiring practices that favor an elite minority of graduates with prestigious institutional affiliations.

One way that some hiring committees are tackling these prestige biases is by providing detailed instructions to applicants on how to prepare their application materials. In theory, detailed applications will allow the hiring committee to focus on evaluating the accomplishments of candidates across common categories rather than ranking based on prestige signals in a CV, such as a name of the applicant's graduate program. This push for equity has resulted in job ads that are often highly prescriptive in the types of documents that applicants should submit. For example, in addition to a cover letter and CV, job ads in many fields now require applicants to submit short statements detailing their previous and future contributions to teaching, research, and diversity. In a comparison of job ads from 1999–2000 and 2019–2020 published in *Anthropology News*, @HelloTristesTropes2021 noticed an increase in the number of materials requested from applicants. For example, twice as many 2019–2020 job ads requested writing samples compared to 1999–2000, and nearly four times as many requested statements of teaching philosophy. In their review of the 'worst job ads of 2021' @dennisWorstAnthroJob2022 reported job ads from Oberlin College and Grinnell College with particularly extreme requirements, which respectively requested nine and thirteen documents from applicants. This high burden on applicants disproportionately favors people with more time and financial resources to prepare the required materials.

A notable requirement that has emerged in recent years is a diversity statement, where the applicant describes their knowledge of, prior contributions to, and future goals for advancing diversity, equity, and inclusion. This requirement has been much-debated largely due to an experiment during 2016–2022 at several University of California campuses that used diversity statements as the first cut for selecting candidates for tenure-track faculty positions [@soucek2021diversity]. Only candidates that scored highly on their diversity statements would have the rest of their application evaluated. This experiment generated intense and widespread public debate about the merits and risks of requiring and using diversity statements. These debates drew attention to the challenges of hiring faculty from underrepresented minorities, and resulted in a variety of approaches to evaluating diversity, even leading some universities to entirely omit requirements for diversity statements in job applications [@guidenAreDiversityStatements].

@@

Our primary data source is the Archaeology Academic Jobs Wiki. Originating in 2007, this is a set of freely accessible web pages that anyone can edit (anonymously or with a free user account) hosted by Fandom, a for-profit company. The Archaeology pages are part of the Academic Jobs Wiki, which coordinates similar collaboratively-edited resources for around 40 academic disciplines. The coordinators and contributors are nearly all anonymous or pseudonymous. Typically, contributors copy and paste all or some of the text of job ads into the wiki, from sources such as the *Chronicle of Higher Education*, *Higher Ed Jobs*, and university websites, collecting ads originally posted in numerous different locations. Other contributors then edit the web page to add comments below an ad to share relevant information based on their experience in applying for that position. These edits result in annotations such as a tally of how many people have applied, the dates of events such as requests for more materials, interviews, offer made, rejection notices, etc. Contributors also edit the page to ask and answer questions about the positions and the application process. These comments make the Academic Jobs Wiki a unique resource for timely and specific information for job-seekers about positions they are interested in, and one of the most important internet resources for the academic job market. Because of its reputation for aggregating ads from diverse sources and rapidly-updated information that is not available elsewhere, the Academic Jobs Wiki has become an authoritative data source for studies of hiring trends in academia [e.g. @Passalacqua_2018; @musial2018five] and a widely recommended resource for applicants [e.g. @lightfoot2021preparing].

@@

Because there is no single editorial authority, collaboratively-edited resources such as the Academic Jobs Wiki can be highly variable in the amount and type of activity over time. It is important to characterise this activity to assess the reliability of the content is as a source of information about the job market. We can get some insights into the dynamics of editing activity from the MediaWiki technology used by Academic Jobs Wiki that creates a public record of every act of adding or removing text to any page in the Wiki [@barrett2008mediawiki]. @fig-panel-wiki-characteristics shows the temporal trends in editing of the Academic Jobs Wiki. Every edit includes the author's identity recorded as either an IP address (a numerical address that includes information about the location of the user's computer) or a username. Of the `r n_all_user_ids` unique editors in our sample, `r per_users_we_can_geolocate`% used IP addresses rather than usernames, and geolocation analysis shows that `r results_ip_addresses_tbl_us_scl`% of IP addresses were based in the US, indicating the Wiki is primarily used by archaeologists in the US. The number of edits and editors roughly halved from 2013 to 2017, and continued to gradually decline. Despite this trend, the size of individual edits slightly increases over time, and the number of edits per job ad is relatively constant over time. The majority of editors are active on the site for less than a year (n = r` edits_for_each_year_tbl_span_tbl_less_than_one_year`, `r edits_for_each_year_tbl_span_tbl_less_than_one_year_perc`%), perhaps reflecting a single season of job-searching. Editors with a longer history likely come from two groups, job seekers participating in multiple seasons of job applications, and faculty on search committees posting multiple job ads.

@@

Individual ads could be recorded to have multiple or none of these topical, geographic, and methods foci, and some of the foci overlap. Some topics include geographic regions because this is how they are typically understood by archaeologists. For example, Mesoamerican archaeology is understood to refer to a specific time period _and_ a specific geographic region. Similarly, we recorded digital archaeology as both a method (when a job ad also had a clearly distinct topical focus, such as historic archaeology) and a topic (when there were no other topics mentioned in the job ad). While this polythetic approach results in categorical overlaps that can make the data challenging to interpret [@kuckartz2014qualitative], in our view this reflects the complex realities of how search committees express their needs when searching for new faculty. Acknowledgement of overlaps also produces new insights into hiring dynamics through revealing intersections between different foci.

The entire R code [@rlanguage] and data files used for all the analyses and visualizations contained in this paper are openly available at https://doi.org/xxx/xx [ for peer reviewers the link is https://zenodo.org/records/14798942?preview=1&token=eyJhbGciOiJIUzUxMiJ9.eyJpZCI6ImMyYWE2MDc3LTMxMzAtNDM1Yi1hYTJkLWMyMGMyNWVhOWMyYiIsImRhdGEiOnt9LCJyYW5kb20iOiJiMjkyMTAyN2UwNzEwNWQ2OWIyZmQ4OTkwMjU0MzY1YyJ9.slnC1BbNgZkiJjrRqmKFK9a1bpI-e9VBbjcU7L5OEeYFs3whcL3POsfVj7phDnoaB5A_6mP9wpGirEaT9Dtvug ] to enable re-use of materials and improve reproducibility and transparency [@Marwick2017]. All of the figures, tables, and statistical test results presented here can be independently reproduced with the code and data in this repository. The code is released under the MIT license, the data as CC-0, and the figures as CC-BY, to enable maximum re-use.

>>>>

Junior scholars with faculty ambitions are thus stuck in a double-bind, forced to funnel their energy into seemingly endless applications for which they receive little to no feedback and which remain invisible on their CVs. As Dennis et al. underscore, early career researchers are "supplicants"—"There is no way for applicants to point out that the job ads are taking too much time out of the scholarly community’s collective time bank" [@dennisWorstAnthroJob2022, pp. 1].Acknowledging the new reality that applying for academic positions is in and of itself a full-time job, a growing number of anthropology graduate programmes are offering professionalization courses that cover the ins and outs of everything from crafting strong cover letters and CVs to performing well in long-list interviews and preparing teaching demonstrations. These courses can provide valuable training in the hidden curriculum of the academy, but the fact that job applications are so complicated that they require training to navigate suggests more disciplinary attention should be paid to the "market" as a historically contingent set of cultural practices. As Rabinow maintains, "if we want ethical considerations to play a central role in the articulation of truth and power—and I think we do—then bringing such considerations into view is the necessary first step towards recognizing who we are today and setting out on the road to a better place" [-@rabinow1992, pp. 71].

@@

PhD graduates who do not go directly into a tenure-track faculty position after graduation often go into a series of short-term appointments as part-time or limited contract instructors, also known as adjunct or contingent faculty. The rising number of contingent faculty has long been a concern in US higher education [@Trevithick2010]. Data from the American Association of University Professors' 2021–22 faculty survey indicate that more than 60% of faculty positions in US universities were held by non-tenure-track full time or part time contingent faculty members [@colby2022annual]. Contingent faculty positions are precarious because they provide low or no health and retirement benefits, limited opportunities for professional development, and lower salaries relative to tenure-track positions. PhD graduates who have a sequence of short-term academic positions are often disadvantaged financially due to low compensation combined with expensive and frequent relocation, restrictive socially due to isolation from family and community, and limited professionally due to being ineligible for many decision-making roles at the universities at which they work [@platzerAcademicPrecarityAmerican2018].

Success in tenure-track job applications in archaeology is strongly determined by where applicants get their PhDs. Data from the 2014–2015 AnthroGuide publication of the American Anthropological Association shows that just ten out of over 100 US graduate programs produced over 30% of the graduates hired into tenure-track faculty positions [@speakman2018choosing]. Top-ranking programs are placing fewer than one in three graduates in tenure-track jobs [@mackieMarketShareAccounting2023]. Similarly, a network analysis of 1,918 faculty holding tenured or tenure-track positions at PhD-granting anthropology programs in the US in 2015 (including 506 archaeologists) found that the just fifteen graduate programs produced 53 percent of tenured and tenure-track positions [@kawaSocialNetworkUS2019]. This network analysis showed that programs with large endowments and widely cited faculty who hold prestigious awards produce the majority of tenured and tenure-track faculty.

Hiring bias predicated on the prestige of specific institutions and programs is not unique to anthropology. Targeted studies of sociology, communication, operations research, and industrial systems engineering show similar dynamics [@barnett2010; @castillo; @feeley2021; @nevin2019]. Broader analyses comparing computer science, business, and history identify comparable patterns among disparate disciplines across the humanities, social sciences, and STEM field—"across disciplines, prestige hierarchies make the most accurate predictions of faculty placement" [@clauset2015, pp. 4]. The most recent comprehensive meta-analysis of faculty placement dynamics in the United States, which examined employment and doctoral education of all tenure-track faculty at PhD-granting universities, underscored the outsize impact of institutional prestige: 80% of faculty trained in the US came from just 20% of institutions [@wapmanQuantifyingHierarchyDynamics2022a]. Within anthropology, such studies draw attention to a central paradox—a discipline purportedly committed to fighting social inequalities continues to reproduce its own systemic inequalities through hiring practices that favor an elite minority of candidates with prestigious affiliations.

One way that some hiring committees are tackling prestige biases is by providing detailed instructions to applicants on how to prepare their application materials. In theory, detailed applications will allow the hiring committee to focus on evaluating the accomplishments of candidates across common categories rather than ranking based on prestige signals in a CV, such as a name of the applicant's graduate program. This push for equity has resulted in job ads that are often highly prescriptive in the types of documents that applicants should submit. For example, in addition to a cover letter and CV, job ads in many fields now require applicants to submit short statements detailing their previous and future contributions to teaching, research, and diversity. In a comparison of job ads from 1999–2000 and 2019–2020 published in *Anthropology News*, @HelloTristesTropes2021 noticed an increase in the number of materials requested from applicants. For example, twice as many 2019–2020 job ads requested writing samples compared to 1999–2000, and nearly four times as many requested statements of teaching philosophy. In their review of the 'worst job ads of 2021' @dennisWorstAnthroJob2022 reported job ads from Oberlin College and Grinnell College with particularly extreme requirements, which respectively requested nine and thirteen documents from applicants. This high burden on applicants disproportionately favors people with more time and financial resources to prepare the required materials.

A notable requirement that has emerged in recent years is a diversity statement, where the applicant describes their knowledge of, prior contributions to, and future goals for advancing diversity, equity, and inclusion. This requirement has been much-debated largely due to an experiment during 2016–2022 at several University of California campuses that used diversity statements as the first cut for selecting candidates for tenure-track faculty positions [@soucek2021diversity]. Only candidates that scored highly on their diversity statements would have the rest of their application evaluated. This experiment generated intense and widespread public debate about the merits and risks of requiring and using diversity statements. These debates drew attention to the challenges of hiring faculty from underrepresented minorities, and resulted in a variety of approaches to evaluating diversity, even leading some universities to entirely omit requirements for diversity statements in job applications [@guidenAreDiversityStatements]. **MORE DETAILS ON THIS HERE—JB WILL ADD.**

@@

Our primary data source is the Archaeology Academic Jobs Wiki. Originating in 2007, the wiki is a set of freely accessible web pages that anyone can edit anonymously or with a free user account. The site is hosted by Fandom, a for-profit company. The Archaeology pages are part of the Academic Jobs Wiki, which coordinates similar collaboratively-edited resources for around 40 academic disciplines. The coordinators and contributors are nearly all anonymous or pseudonymous. Typically, contributors copy and paste all or some of the text of job ads into the wiki, from a variety of sources such as the *Chronicle of Higher Education*, *Higher Ed Jobs*, and university websites. Other contributors then edit the web page to add comments below an ad to share relevant information based on their experience in applying for that position. These edits result in annotations such as a tally of how many people have applied, the dates of events such as requests for more materials, interviews, offer made, rejection notices, etc. Contributors also edit the page to ask and answer questions about the positions and the application process. These comments have made <!--If we are a bit more slippery about our tenses, we don't need to argue that the wiki is STILL crucial, esp. if late we will discuss its decline--> the Academic Jobs Wiki a unique resource for timely and specific information for job-seekers about positions they are interested in, and one of the most important internet resources for the academic job market. Because of its reputation for aggregating ads from diverse sources and rapidly-updated information that is not available elsewhere, the Academic Jobs Wiki has become an authoritative data source for studies of hiring trends in academia [e.g. @Passalacqua_2018; @musial2018five] and a widely recommended resource for applicants [e.g. @lightfoot2021preparing].

@@

Because there is no single editorial authority, collaboratively-edited resources such as the Academic Jobs Wiki can be highly variable in the amount and type of activity over time. It is important to characterise this activity to assess the reliability of the content is as a source of information about the job market. We can get some insights into the dynamics of editing activity from the MediaWiki technology used by Academic Jobs Wiki that creates a public record of every act of adding or removing text to any page in the Wiki [@barrett2008mediawiki]. @fig-panel-wiki-characteristics shows the temporal trends in editing of the Academic Jobs Wiki. Every edit includes the author's identity recorded as either an IP address (a numerical address that includes information about the location of the user's computer) or a username. Of the `r n_all_user_ids` unique editors in our sample, `r per_users_we_can_geolocate`% used IP addresses rather than usernames, and geolocation analysis shows that `r results_ip_addresses_tbl_us_scl`% of IP addresses were based in the US, indicating the Wiki is primarily used by archaeologists in the US. The number of edits and editors roughly halved from 2013 to 2017, and continued to gradually decline. Despite this trend, the size of individual edits slightly increases over time, and the number of edits per job ad is relatively constant over time. The majority of editors are active on the site for less than a year (n = r`edits_for_each_year_tbl_span_tbl_less_than_one_year`, `r edits_for_each_year_tbl_span_tbl_less_than_one_year_perc`%), perhaps reflecting a single season of job-searching. Editors with a longer history likely come from two groups, job seekers participating in multiple seasons of job applications, and faculty on search committees posting multiple job ads.

@@

Individual ads could be recorded to have multiple or none of these topical, geographic, and methods foci, and some of the foci overlap. Some topics include geographic regions because this is how they are typically understood by archaeologists. For example, Mesoamerican archaeology is understood to refer to a specific time period *and* a specific geographic region. Similarly, we recorded digital archaeology as both a method (when a job ad also had a clearly distinct topical focus, such as historic archaeology) and a topic (when there were no other topics mentioned in the job ad). While this polythetic approach results in categorical overlaps that can make the data challenging to interpret [@kuckartz2014qualitative], in our view this reflects the complex realities of how search committees express their needs when searching for new faculty. Acknowledgement of overlaps also produces new insights into hiring dynamics through revealing intersections between different foci.

The entire R code [@rlanguage] and data files used for all the analyses and visualizations contained in this paper are openly available at https://doi.org/xxx/xx \[ for peer reviewers the link is https://zenodo.org/records/14798942?preview=1&token=eyJhbGciOiJIUzUxMiJ9.eyJpZCI6ImMyYWE2MDc3LTMxMzAtNDM1Yi1hYTJkLWMyMGMyNWVhOWMyYiIsImRhdGEiOnt9LCJyYW5kb20iOiJiMjkyMTAyN2UwNzEwNWQ2OWIyZmQ4OTkwMjU0MzY1YyJ9.slnC1BbNgZkiJjrRqmKFK9a1bpI-e9VBbjcU7L5OEeYFs3whcL3POsfVj7phDnoaB5A_6mP9wpGirEaT9Dtvug \] to enable re-use of materials and improve reproducibility and transparency [@Marwick2017]. All of the figures, tables, and statistical test results presented here can be independently reproduced with the code and data in this repository. The code is released under the MIT license, the data as CC-0, and the figures as CC-BY, to enable maximum re-use.
